## Dataset

A dataset on red wine quality was downloaded from https://archive.ics.uci.edu/dataset/186/wine+quality. The dataset was originally published in "Modeling wine preferences by data mining from physicochemical properties," by By P. Cortez, A. Cerdeira, Fernando Almeida, Telmo Matos, J. Reis. 2009. Decision Support Systems. Though the original paper examined both red and white wine variants of the Portuguese "Vinho Verde" wine, for this project I will only consider the red wine dataset.

The dataset contains the following variables:

fixed.acidity - tartaric acid ($g/dm^3$)

volatile.acidity - acetic acid ($g/dm^3$)

citric.acid ($g/dm^3$)

residual.sugar ($g/dm^3$)

chlorides - sodium chloride ($g/dm^3$)

free.sulfur.dioxide ($mg/dm^3$)

total.sulfur.dioxide ($mg/dm^3$)

density ($g/cm^3$)

pH

sulphates - potassium sulfphate ($g/dm^3$)

alcohol (% vol.)

quality - a subjective score between 0 and 10 based on sensory data

Due to privacy concerns, potentially identifying information about the wine, such as grape variety, region grown, brand, or price, are not available.

## Getting to know the variables

```{r}
wine <- read.csv("Data/winequality-red.csv") 
summary(wine) 
library(tidyverse) 
hist(wine$quality, breaks=c(0.5:9.5)) 
mean(wine$quality) 
sd(wine$quality) 
max(wine$quality) 
min(wine$quality) 
median(wine$quality) 
IQR(wine$quality)  
nrow(wine[wine$quality >4 & wine$quality <7,])/nrow(wine)   
library(reshape2) 
boxwine <- melt(wine) 
library(ggplot2) 
ggplot(boxwine, aes(factor(variable),value)) +   
  geom_boxplot() +   
  facet_wrap(~variable,scale="free", ncol=4)  
pairs(wine) 
```

Looking at distribution of each variable and its correlation with the outcome variable, quality.

#### Fixed Acidity

```{r}
hist(wine$fixed.acidity) 
ggplot(aes(x=fixed.acidity, y=quality), data=wine) + geom_point()
```

fixed.acidity is roughly normally distributed with some right skew. It does not appear to have any relationship with quality.

#### Volatile Acidity

```{r}
hist(wine$volatile.acidity)  
ggplot(aes(x=volatile.acidity, y=quality), data=wine) + geom_point()
```

volatile.acidity is roughly normally distributed with some right skew. There may be a slight negative correlation between the two variables, but it is likely not linear - I might include cubic terms here.

#### Citric Acid

```{r}
hist(wine$citric.acid)  
ggplot(aes(x=citric.acid, y=quality), data=wine) + geom_point()
```

citric.acid is NOT normally distributed - many wines have very low amounts of citric acid, so it is bounded by 0 $g/dm^3$ on the left. There is no apparent relationship between citric.acid and quality.

#### Residual Sugar

```{r}
hist(wine$residual.sugar)  
ggplot(aes(x=residual.sugar, y=quality), data=wine) + geom_point()
```

residual.sugar has a very positively skewed distribution with no clear relationship to quality.

#### Chlorides

```{r}
hist(wine$chlorides)  
ggplot(aes(x=chlorides, y=quality), data=wine) + geom_point()
```

chlorides is extremely leptokurtic with positive skew. Wines with higher concentrations of chlorides appear to have been rated as lower quality, but the relationship may not be linear - I would include cubic terms.

#### Free Sulfur Dioxide

```{r}
hist(wine$free.sulfur.dioxide)  
ggplot(aes(x=free.sulfur.dioxide, y=quality), data=wine) + geom_point()
```

free.sulfur.dioxide has positive skew, with very little bearing on wine qualitiy.

#### Total Sulfur Dioxide

```{r}
hist(wine$total.sulfur.dioxide)  
ggplot(aes(x=total.sulfur.dioxide, y=quality), data=wine) + geom_point()
```

fixed.acidity is roughly normally distributed with some right skew. It does not appear to have any relationship with quality.

#### Density

```{r}
hist(wine$density)  
ggplot(aes(x=density, y=quality), data=wine) + geom_point()
```

fixed.acidity is roughly normally distributed with some right skew. It does not appear to have any relationship with quality.

#### pH

```{r}
hist(wine$pH)  
ggplot(aes(x=pH, y=quality), data=wine) + geom_point()
```

fixed.acidity is roughly normally distributed with some right skew. It does not appear to have any relationship with quality.

#### Sulphates

```{r}
hist(wine$sulphates)  
ggplot(aes(x=sulphates, y=quality), data=wine) + geom_point()
```

fixed.acidity is roughly normally distributed with some right skew. It does not appear to have any relationship with quality.

#### Alcohol

```{r}
hist(wine$alcohol)  
ggplot(aes(x=alcohol, y=quality), data=wine) + geom_point()
```

fixed.acidity is roughly normally distributed with some right skew. It does not appear to have any relationship with quality.

## The Model

Here's the maximum model including all predictor variables:

```{r}
library(foreign)
library(MASS)
mod_ord <- polr(as.factor(quality) ~ .*., wine,Hess=TRUE)
summary(mod_ord)

mod_ord_best <- step(mod_ord, direction="both",trace=0)
summary(mod_ord_best)
vif(mod_ord_best)

# Not able to proceed with NaNs included - seem to be induced by free.sulfur.dioxide:total.sulfur.dioxide, so that interaction needs to be removed from the model and rerun
rm_var <- "free.sulfur.dioxide:total.sulfur.dioxide"
mod_ord_nextbest <- update(mod_ord_best,paste0(".~. - ", rm_var))
summary(mod_ord_nextbest)
vif(mod_ord_nextbest)

# Ordinarily I'd used a max VIF of 10, but with that there are SIGNIFICANT multicollinearity problems left in the model (between density and alcohol, for instance, which is practically a linear function on its own), so I set a ceiling of 5 instead
while(max(vif(mod_ord_nextbest))>5){
  rm_var <- max_vif(mod_ord_nextbest)
  mod_ord_nextbest <- update(mod_ord_nextbest,paste0(".~. - ", rm_var))
}
summary(mod_ord_nextbest)
vif(mod_ord_nextbest)
pairs(profile(mod_ord_nextbest)) # visualizes which pairs of terms are multicollinear

library(PResiduals)
pres <- presid(mod_ord_nextbest)

library(sure)
library(ggplot2)
p1 <- ggplot(data.frame(x=resids(mod_ord_nextbest), y=pres), aes(x,y)) +
  geom_point(color="#444444",shape=19,size=2,alpha=0.5) +
  geom_smooth(color="red",se=FALSE) +
  ylab("Probability-scale residual")

p2 <- ggplot(data.frame(y=pres),aes(sample=y)) +
  stat_qq(distribution=qunif,dparams=list(min=-1,max=1),alpha=0.5) +
  geom_qq_line(distribution=qunif,dparams=list(min=-1,max=1)) + 
  xlab("Sample quantile") +
  ylab("Theoretical quantile")

p1
p2


# Create null model to run the Chi-square test
mod_ord_null <- polr(as.factor(quality) ~ 1, wine,Hess=TRUE)


summary(mod_ord_nextbest)

confint(profile(mod_ord_nextbest)) # 95% CI for betas
exp(confint(profile(mod_ord_nextbest))) # 95% CI for betas

X <- summary(mod_ord_null)$deviance - summary(mod_ord_nextbest)$deviance
p <- pchisq(X,df=9,lower.tail=FALSE)
c(X,p)

rsq <- (summary(mod_ord_null)$deviance - summary(mod_ord_nextbest)$deviance)/summary(mod_ord_null)$deviance
rsq

```
